### 2013-09-17 thesis-proposal/background.tex
Updated abstract, most of introduction, beginning of approach
updated proposal
Updated beginning of proposal

\section{Background}
We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.

\subsection{Language-Oriented Approaches}
\subsubsection{Language Frameworks}
A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review). In some cases, these tools allow language features to be defined modularly and composed differentially to produce a variety of different languages. To our knowledge, none of these mechanisms guarantee that any combination of features can be safely composed, nor do they guarantee safety properties about the resulting languages. User-defined code is still ultimately compiled against a particular language, and because language composition cannot be automatic or guaranteed safe, interoperability with code written in other languages is thus limited by the difficulties described above despite the modular construction.

\subsubsection{Extensible Compilers and Tools}
A related methodology is to implement language features as compiler and tool extensions directly. A number of extensible compilers have been developed to support this approach (see \cite{clements2008comparison}). As with language frameworks, this approach can lead to \emph{de facto} implementation of a new language, since user libraries must be compiled against a particular combination of extensions, and these extensions cannot be guaranteed to compose in general. Moreover, a language may have multiple competing compilers and other tools. By relying on implementation-specific features of a single tool to define core semantics and behaviors, the clean conceptual separation between languages and tools is broken, leading to compatibility issues and hard-to-anticipate behaviors for user code. We argue that this approach should be considered harmful.

\subsection{Library-Oriented Approaches}
\subsubsection{Embedded DSLs}
Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language. In languages with rich type systems, such as Haskell, this approach can be quite successful (e.g. \cite{svensson2011obsidian}). Ultimately, however, this approach is limited by the host system, and as discussed in Section 2, thus limits experts who want to express particularly novel constructs, relative to the host language.

\subsubsection{Term Rewriting Systems}
Many languages and tools allow developers to rewrite expressions according to custom rules. These can broadly be classified as {\it term rewriting systems}. Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example. Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g. MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages. These systems are expressive if used correctly, but verifying correctness and non-interference properties is difficult for the same reason. Manipulating source trees directly is a complex task, even in languages with simple grammars like LISP. Finally, term rewriting systems focus on rewriting terms to support alternative language semantics but do not intrinsically support extensions that cover the full programming system.

\subsubsection{Active Libraries}
Active libraries, as proposed by Czarnecki et al. \cite{activelibraries}, ``are not passive collections of routines or objects, as are traditional libraries, but take an active role in generating code''. They go on to suggest a number of areas in which the library could interact with the programming system, including optimizing code, checking source code for correctness, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations''.

This paper was largely a proposal. The concrete implementations of this concept have largely been term rewriting systems described above. One prominent example within the active libraries literature is Blitz++, which uses the C++ template expansion system as a metalanguage to support optimizations of array operations. An example of tool support comes from the Tau package for tuning and analyzing parallel programs. Tau allows libraries to instrument themselves declaratively to hide internal details and complex internal representations from users. A more extensive system with support for active libraries is Xroma. Xroma allows users to provide annotations that intercept compilation of a component at various stages of compilation to support rewriting, custom error handling and custom error checking. The approach taken by Xroma, by still requiring direct syntax manipulation and allowing arbitrary interception, is flexible but not compositional, and generally more complex than may be necessary. There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.

>>>>

% !TEX root = omar-thesis-proposal.tex
\subsubsection{Background: Active Libraries}\label{alibs}
The term \emph{active libraries} was first used by Veldhuizen et al. \cite{atthesis, atpaper} to describe ``libraries that take an active role in compilation, rather than being passive collections of subroutines''. The authors went on to suggest a number of concrete reasons libraries might benefit by being able to influence the programming system at compile-time or edit-time, including high-level program optimization, checking programs for correctness against specialized criteria, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations'' (anticipating interactions between libraries and tools other than just the compiler. Our definition, given in Section \ref{motivation}, differs  from theirs, partly for this reason).

% This stuff isn't actually hooking into the compiler...
The first concrete realizations of active libraries, prompting the introduction of the term, were libraries that performed domain-specific program optimization at compile-time by exploiting language mechanisms that allow for limited compile-time computation. A prominent example in the literature is Blitz++, a library that uses C++ template metaprogramming to optimize compound operations on vectors and matrices by eliminating intermediate allocations \cite{blitzpp}. Although this and several other interesting optimizations have been shown possible by this technique, its expressiveness is fundamentally limited because template expansion allow for only the substitution of compile-time constants into pre-written code, and template metaprograms are notoriously difficult to read, write, debug and reason about. %They are, however, composable, because templates are associated with a single function or class, and  the only metatheoretic issue is that template expansion can be non-terminating.

More powerful compile-time \emph{term rewriting mechanisms} integrated into some languages can also be used for optimization, as well as for inserting specialized error checking and reporting\- logic and extending the language with powerful domain-specific abstractions. These mechanisms are highly expressive because they allow users to programmatically manipulate syntax trees directly, but they suffer from problems of composability and safety. Macros, such as those in MetaML \cite{metaml}, Template Haskell \cite{template_haskell} and others\todo{reference other macro systems}, can take full control over all the code that they enclose, so there is no way to guarantee that nested macros will compose as intended. Outer macros can neglect to invoke or  manipulate the output of inner macros in ways that can cause conflicts or weaken important guarantees. Once data escapes a macro's scope, there is no way to rely on the guarantees and features that were available within its scope -- the output of a macro is a value in the underlying language, so the underlying language is all that governs it (a problem related to the interoperability problem of Section \ref{motivation}). It can also be difficult to reason about the semantics of code when a number of enclosing macros may be manipulating it.

Some rewriting systems go beyond the block scoping of macros. Xroma (pronounced ``Chroma''), for example, is designed around active libraries and allows users to insert custom rewriting and error checking passes into the compiler from within libraries \cite{xroma}. Similarly, the derivative of Haskell implemented by the Glasgow Haskell Compiler (GHC) allows libraries to define custom compile-time term rewriting logic if an appropriate flag is passed in \cite{haskell-rewrite}. In both cases, the user-defined logic can dispatch on arbitrary patterns of code throughout the component or program the extension is active in, so these mechanisms are highly expressive and avoid some of the difficulties of block-scoped macros. But libraries containing such global rewriting logic cannot be safely composed because two different libraries may attempt to rewrite the same piece of code differently. It is also difficult to guarantee that such logic is correct and difficult for users to reason about code when simply importing a library can change the semantics of the program in a non-local manner.
% http://www.haskell.org/ghc/docs/latest/html/users_guide/rewrite-rules.html

Another example of an active library approach to extensibility is SugarJ \cite{sugarj}. SugarJ allows libraries to extend the base syntax of Java in a nearly-arbitrary manner, and these extensions are imported transitively throughout a program. SugarJ libraries are thus also not safely composable. For example, a library that defines a literal syntax for HTML would conflict with another that defines a literal syntax for XML because they define differing semantics for some of the same syntactic forms. If SugarJ was used by two different regular expression engines to provide literal syntax for regular expression patterns, there could easily be conflicts at link-time because both will introduce many of the same standard notations but back them with differing implementations. 
And again, it can be difficult for users to predict what type of data an unfamiliar piece of syntax desugars into, leading to difficulties reading and reasoning about code.

%Another example comes from the Tau package for tuning and analyzing parallel programs, which allows libraries to instrument themselves declaratively so that internal details can be hidden from users of the tool \cite{tau}.

%The researchers behind these tools often claim that such conflicts are rare. There have, to our knowledge, been no studies of this claim. Nevertheless, the core fact remains: conflicts can arise, which makes it difficult to rely on such libraries

%There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.

%\todo{Smooth this transition}Developers specify type-specific compile-time and edit-time features by equipping type families, when they are defined, with functions written in an appropriate metalanguage. These are selectively invoked by tools, such as the parser, type checker, translator and editor, when they work with expressions of that type, and only in those situations.

%\subsection{Example: First-Class Natural Numbers}
%To make our approach more concrete and emphasize its foundational character, let us begin with a very simple example, first-class natural numbers. If a programming system included first-class support for natural numbers, it might support behaviors like the following:
%
%\begin{enumerate}
%\item type checking rules that constrain uses of the three operators associated with the $\nat$ type, as defined in G\"{o}del's System \textbf{T} \cite{tapl}: $\natz$, $\nats{e}$ and $\natrec{e_1}{e_2}{e_3}$\todo{phrasing of his}
%\item type checker error messages that provide domain-specific assistance, like detecting when the trailing $\textbf{z}$ in a constant has been omitted: \todo{phrasing / merge this into 1?}
%\begin{verbatim}
%    s(s(s))
%        ^ (Line 1, Character 5)
%    Error: Operator s requires 1 argument of type nat. Provided 0 arguments.
%    Did you mean s(z)?
%\end{verbatim}
%\item translation into an efficient integer representation at run-time
%\item syntax support \todo{elaborate on syntax support for nat literals}
%\item editor support, such as the ability to instantiate nat constants more easily \todo{elaborate on this / reference Graphite section}
%%generate statistical predictions useful for code completion (e.g. biasing smaller constants) and other editor interactions, customized syntactic/visual representations and so on.
%\end{enumerate}
%

%\section{Related Work}
%We review existing approaches that are available to researchers and domain experts who wish to develop novel constructs below.
%
%\subsection{Language-Oriented Approaches}
%\subsubsection{Language Frameworks}
%A number of tools have been developed to assist with this task of developing new languages and their associated tools, including compiler generators, language workbenches and domain-specific language frameworks (see \cite{fowler2010domain} for a review). In some cases, these tools allow language features to be defined modularly and composed differentially to produce a variety of different languages. To our knowledge, none of these mechanisms guarantee that any combination of features can be safely composed, nor do they guarantee safety properties about the resulting languages. User-defined code is still ultimately compiled against a particular language, and because language composition cannot be automatic or guaranteed safe, interoperability with code written in other languages is thus limited by the difficulties described above despite the modular construction.
%
%\subsubsection{Extensible Compilers and Tools}
%A related methodology is to implement language features as compiler and tool extensions directly. A number of extensible compilers have been developed to support this approach (see \cite{clements2008comparison}). As with language frameworks, this approach can lead to \emph{de facto} implementation of a new language, since user libraries must be compiled against a particular combination of extensions, and these extensions cannot be guaranteed to compose in general. Moreover, a language may have multiple competing compilers and other tools. By relying on implementation-specific features of a single tool to define core semantics and behaviors, the clean conceptual separation between languages and tools is broken, leading to compatibility issues and hard-to-anticipate behaviors for user code. We argue that this approach should be considered harmful.
%
%\subsection{Library-Oriented Approaches}
%\subsubsection{Embedded DSLs}
%Embedded domain-specific languages are languages that creatively repurpose existing language constructs to create interfaces that resemble those of a distinct language. In languages with rich type systems, such as Haskell, this approach can be quite successful (e.g. \cite{svensson2011obsidian}). Ultimately, however, this approach is limited by the host system, and as discussed in Section 2, thus limits experts who want to express particularly novel constructs, relative to the host language. 
%
%\subsubsection{Term Rewriting Systems}
%Many languages and tools allow developers to rewrite expressions according to custom rules. These can broadly be classified as {\it term rewriting systems}. Macro systems, such as those characteristic of the LISP family of languages \cite{mccarthy1978history}, are the most prominent example. Some compile-time metaprogramming systems also allow users to manipulate syntax trees (e.g. MetaML \cite{Sheard:1999:UMS}), and external rewrite systems also exist for many languages. These systems are expressive if used correctly, but verifying correctness and non-interference properties is difficult for the same reason. Manipulating source trees directly is a complex task, even in languages with simple grammars like LISP. Finally, term rewriting systems focus on rewriting terms to support alternative language semantics but do not intrinsically support extensions that cover the full programming system.
%
%\subsubsection{Active Libraries}
%Active libraries, as proposed by Czarnecki et al. \cite{activelibraries}, ``are not passive collections of routines or objects, as are traditional libraries, but take an active role in generating code''. They go on to suggest a number of areas in which the library could interact with the programming system, including optimizing code, checking source code for correctness, reporting domain-specific errors and warnings, and ``rendering domain-specific textual and non-textual program representations and for interacting with such representations''. 
%
%This paper was largely a proposal. The concrete implementations of this concept have largely been term rewriting systems described above. One prominent example within the active libraries literature is Blitz++, which uses the C++ template expansion system as a metalanguage to support optimizations of array operations. An example of tool support comes from the Tau package for tuning and analyzing parallel programs. Tau allows libraries to instrument themselves declaratively to hide internal details and complex internal representations from users. A more extensive system with support for active libraries is Xroma. Xroma allows users to provide annotations that intercept compilation of a component at various stages of compilation to support rewriting, custom error handling and custom error checking. The approach taken by Xroma, by still requiring direct syntax manipulation and allowing arbitrary interception, is flexible but not compositional, and generally more complex than may be necessary. There also does not appear to be a clear, well-founded theoretical foundation for this approach, nor for active libraries in general.

### 2013-09-17 thesis-proposal/conclusion.tex
updated proposal
Updated beginning of proposal

Monolithic programming systems enforce a dichotomy between \emph{built-in constructs}, which enjoy support throughout the programming system but must be designed and implemented by the system designer in advance, and \emph{user-defined constructs}, which can be distributed as user libraries, but must creatively combine and repurpose the small set of available built-in constructs to express all desired run-time, compile-time and edit-time behaviors. This dichotomy has forced researchers and domain experts who want to significantly advance the state-of-the-art in programming systems, particularly in compiled, statically-typed systems where this dichotomy is most salient, to design and develop new languages. But this then couples their core innovations with a collection of unrelated design decisions, requires more effort both for the experts developing the innovation and its targeted user community, and leads to intrinsic problems for users developing applications consisting of components written in different languages.
\newpage
\listoftodos
\newpage

>>>>

\newpage
\listoftodos
\newpage

### 2013-09-17 thesis-proposal/problem.tex
Updated beginning of proposal

% !TEX root = omar-thesis-proposal.tex
\section{Motivation}\label{motivation}\begin{figure}
\begin{center}
\includegraphics[scale=0.45]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-external approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages if the interface uses common constructs and an interoperability layer has been developed (the \emph{interoperability problem}). (b) With the language-internal approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries. If the extension mechanism guarantees the safety of arbitrary compositions of extensions, the necessary primitives can simply be imported by library clients as needed, and so interoperability is not a problem.}
\label{approaches}
\end{figure}
Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. It has become clear that no small, fixed collection of primitive abstractions and tools can fully satisfy these criteria in all situations. Rather, researchers and domain experts must continue, in a principled manner, to design and implement novel abstractions, provide alternative implementations of existing abstractions, and supplement languages with new tools, and tools with new behaviors, to better satisfy these criteria within the constraints of different problem domains.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.

When possible, taking a \emph{language-internal approach} to implementing a new abstraction or system behavior (collectively, a new \emph{feature}) is simpler and more practical. If a feature can be realized by creatively repurposing existing features and distributed as a library, both providers and clients face fewer barriers to adoption because it is easy to gradually integrate library-based features into existing code and they leverage well-understood and well-developed mechanisms. But taking this approach is often \emph{not} possible today because libraries are generally vehicles for specifying the run-time behavior of a program. The compile-time and edit-time behaviors of a programming system cannot easily be adapted or extended in a specialized way from within libraries in most programming systems. That is, the language's syntax, static semantics and corresponding dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and the other tools, like code editors and debuggers, operate according to fixed, domain-agnostic protocols based on these fixed semantics.

If any of these components of the system must be modified to realize a new feature, providers must instead take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language} \cite{dsl}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler or other tool. %Extending a particular tool is morally equivalent, for our purposes, to creating a new system, because we aim to minimize the extent to which the edit-time, compile-time and run-time semantics of libraries depend on the implementation details of a particular tools.

Unfortunately, when providers of new features take language-external approaches, it causes problems for clients. Features become coupled to a collection of other unrelated and often underdeveloped features, making adoption more costly when these other choices are not appropriate. For example, although recent evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations remain more widely adopted because they do not require porting applications to a language designed around a few privileged parallel programming abstractions. These abstractions may only be useful in performance-critical portions of the application, and then only in cases where no competing abstraction would be more appropriate.

@@

%Clients seem to prioritize the ability to choose different features for different portions of an application. 
If it were easy to call between languages, then having a variety of specialized languages would be less of a problem. In fact, some have advocated for just this development model, calling it the the \emph{language-oriented approach} to software development \cite{languageoriented}. A fundamental problem with this approach, however, arises at the interfaces between languages. The specialized abstractions particular to one language cannot be safely and naturally expressed in another, in general, so building a program out of components written in many different languages is difficult or impossible whenever these specialized abstractions are exposed at library interface boundaries. We refer to this as the \emph{interoperability problem}.

One strategy increasingly being taken by programming language designers to partially address the interoperability problem today is to target an established language, such as the Java Virtual Machine (JVM) bytecode, and support a superset of its constructs. Scala \cite{scala} and F\# \cite{fsharp} are examples of general-purpose languages that have taken this approach. This only provides full interoperability in one direction (Figure \ref{approaches}a). While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the constructs available in the common language. If full interoperability is desired, new languages must only include constructs that can already be expressed safely and reasonably naturally in the common language. Many innovative features, however, can be difficult to define in terms of existing features in ways that guarantee all necessary invariants are maintained and that do not require large amounts of boilerplate code. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this domain-specific invariant is left completely unchecked at language boundaries. In Scala, interfaces built around traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. In some cases, desirable features must be omitted entirely due to concerns about  interoperability. The module system in F\#, for example, is substantially simpler than that in its predecessor, OCaml, despite F\# otherwise aiming to maintain compatibility, due to the need for bidirectional interoperability between F\# and other CLI languages.
%\end{itemize}

For these reasons, we argue that the language-oriented approach is fundamentally flawed and so taking a language-external approach to realizing a new feature, where new features are packaged into mutually incompatible languages and associated tools, should be considered harmful and avoided when possible. The goal of the research proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been centrally-controlled by  the  committees that govern languages or tools\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- languages more closely resemble modern market economies.}\todo{read Arch D. Robison. Impact of economics on compiler optimization.}. Specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{translation} (the first stage of compilation) and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. Such libraries are called \emph{active libraries} because, rather than being passive consumers of features already available in the system, they contain logic invoked directly by tools to provide new domain-specific functionality during development or compilation \cite{active-libraries}. Features implemented within active libraries can be imported as needed by clients of libraries that rely on them, unlike features implemented by language-external means.

Some critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given to users, the system may become unreliable. For example, an extension could weaken important metatheoretic guarantees previously provided by the system. Type safety, for instance, may no longer hold if the static and dynamic semantics of the language can be modified arbitrarily within libraries. Furthermore\-, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could lead to subtle problems at link-time. For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program. These kinds of safety issues have plagued previous attempts at language-internal extensibility (discussed further in the next section). To prevent them, the system must strictly constrain extensions such that they are guaranteed safe in isolation and also safely composable in any combination.

The mechanisms described in this thesis are designed to be highly \textbf{expressive}, allowing library-based implementations of features comparable to built-in features found in modern programming systems, without compromising on these important notions of \textbf{safety}. This is accomplished by organizing extension logic around types, rather than languages. By scoping new compile-time and edit-time logic to expressions of the type with which it is associated, rather than applying it globally or only within contiguous syntactic blocks as in many previously-developed mechanisms, conflicts are avoided.  We call types with such logic associated with them \emph{active types} and systems that support them \emph{actively-typed programming systems}. By constraining the extension logic with a type system and using techniques from the compiler correctness literature, we can ensure that the system as a whole maintains important safety properties. %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.

>>>>

% !TEX root = omar-thesis-proposal.tex
\vspace{-25pt}
\section{Motivation}\label{motivation}
Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, helps users identify and fix errors as early as possible, supports a natural programming style, and that performs well across diverse problem domains and hardware platforms remains a grand challenge in computing. Due to the increasing diversity and complexity of modern problem domains and hardware platforms, it has become clear that no small  collection of general-purpose primitives backed by a simple toolchain will be able to fully satisfy these criteria in all cases. Instead,  researchers will continue to provide specialized abstractions, notations,  type systems, implementation strategies, optimizations, run-time systems and tool support (collectively, specialized \emph{features}) in order to meet these growing challenges. %, researchers and domain experts must continue to design and develop, in a principled manner, specialized notations, type systems, optimizations, run-time systems and tools (collectively, \emph{features}) to support developers as they work in increasingly complex problem domains and on modern hardware systems.

% Rather, researchers and domain experts must continue, in a principled manner, to specify and implement novel abstractions, provide alternative implementations of existing abstractions, supplement languages with useful tools and improve tools with new behaviors (collectively, realize new \emph{features}) that better satisfy these criteria by balancing the concerns relevant to particular application domains.

Ideally, providers would develop and distribute new features as libraries, so that developers could granularly choose those that best satisfy the constraints of their problem domains. Unfortunately, this is often infeasible, because libraries are nearly exclusively vehicles for specifying the run-time behaviors of a program, but providing these kinds of features  requires or benefits from some measure of control over the compile-time or edit-time semantics of the programming system. That is, from the perspective of a library, the language's semantics are fixed in advance, the compiler and run-time system are ``black box'' implementations of these fixed semantics, and the other tools, like code editors and debuggers, operate according to domain-agnostic protocols also based exclusively on these fixed semantics. As a result, providers of new system features must take \emph{language-external approaches}. We will argue that such approaches have several fundamental problems, and that taking them has led to an unnecessary gap between research and practice. In place of these approaches, we will advocate for \emph{language-internal extensibility mechanisms}, and show how, by organizing new features around types and constraining them appropriately, such mechanisms can be made both safe and highly expressive.
%But taking a \emph{language-internal approach} to implementing a feature is the most practical. If a feature can be realized by creatively using existing language constructs and distributed as a library, clients face fewer barriers to adoption because it is easy to integrate library-based features into existing projects gradually and granularly and they leverage well-understood and well-developed mechanisms.
% But taking this approach is often \emph{not} possible today %We call designs \emph{monolithic programming systems}.

%To realize a new abstraction or system behavior, such experts can consider either a \emph{language-internal approach}, where they work within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they create a new, distinct programming system (often centered around what has come to be called a new \emph{domain-specific language} \cite{dsl}) or extend an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler, editor or other tool.
\subsection{Motivating Example: Regular Expressions}\label{regex}
To make the issue concrete, let us begin with a simple example. \emph{Regular expressions} are a widely-used mechanism for finding patterns in unstructured and semi-structured strings (when formal grammars are not appropriate) \cite{regex}. If a programming system included full compile-time and edit-time support for regular expressions, it might provide features like:

\begin{enumerate}
\item \textbf{built-in syntax for pattern literals} so that malformed patterns result in intelligible \emph{compile-time} parsing errors, motivated by the frequency of run-time errors relating to pattern syntax in Java \cite{regex-type-system} and other languages that do not have this feature.
\item \textbf{type checking logic} that ensures that key constraints related to regular expressions are not violated, such as that out-of-bounds group indices are not used \cite{regex-type-system} and that only values with correct types are spliced into regular expressions. When a type error is found, the error message should again be intelligible.
\item \textbf{translation logic} that partially or fully compiles regular expressions into the efficient internal representation that will be used by the regular expression matching engine at run-time. In most languages, this compilation step occurs at run-time, even if the pattern is fully known at compile-time, thereby introducing latency into programs. If the developer is not careful, regular expressions used repeatedly in a program might be needlessly re-compiled on each use. %By performing this step ahead-of-time, these dangers can be avoided.
\item \textbf{editor-integrated tooling} for interactively testing patterns against example strings, quickly referring to documentation, searching databases of common patterns and other domain-specific edit-time facilities.
\end{enumerate}

%In a conventional \emph{monolithic} programming system, support for each of these features would need to be built into the language and tools. 
No system today has built-in support for all of the features enumerated above. Instead, libraries generally provide support for regular expressions by leveraging  general-purpose constructs. Unfortunately, it is impossible to fully encode the syntax and the specialized static and dynamic semantics described above in terms of standard general-purpose notations and abstractions. Library providers have thus needed to compromise, typically by representing regular expressions using strings and deferring parsing, typechecking and translation to run-time, which introduces performance overhead and can lead to unanticipated run-time errors (as shown in \cite{regex-type-system}) and security vulnerabilities (due to injection attacks when splicing is used, for example). Similarly, edit-time features, as described above, are rarely integrated into editors, and never in a way that facilitates their discovery and use directly when the developer is manipulating regular expressions. In most cases, tools must be discovered independently and accessed externally (for example, via a browser), making both their development and use less common and more awkward than necessary \cite{active-code-completion} (see Section \ref{acc}).

\subsection{Language-External Approaches}
\begin{figure}
\begin{center}
\includegraphics[scale=0.45]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-external approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages if the interface uses common constructs and an interoperability layer has been developed (the \emph{interoperability problem}). (b) With the active library approach, there is one extensible host language and the compile-time and edit-time logic governing novel constructs is expressed within libraries. If the extension mechanism guarantees the safety of arbitrary compositions of extensions, the necessary primitives can simply be imported by library clients as needed, and so interoperability is not a problem.}
\label{approaches}
\end{figure}

When the compile-time or edit-time semantics of a system must be modified to fully realize a new feature, as in the example above, providers often take a \emph{language-external approach}, either by developing a new or derivative programming system (often centered around a so-called \emph{domain-specific language} \cite{dsl}) or by extending an existing system by some mechanism that is not part of the language itself, such as an extension mechanism supported by a {particular} compiler\footnote{Note that compilers that modify or allow modification of  the semantics of their base language, rather than simply implementing meaning-preserving optimizations, should be considered as pernicious means for creating new languages. Many compilers, including \texttt{gcc}, \texttt{GHC} and \texttt{SML/NJ}, are guilty of this sin, meaning that some programs that seem to be written in C, Haskell or Standard ML are actually written in tool-specific derivatives of those languages. Language-internal mechanisms do not lead to such fragmentation.} or other tool. %This latter method couples the semantics of the feature to the implementation details of a particular tool. Because the use of one implementation entails a different semantics for the feature than another, the extended tool acts, \emph{de facto}, as a distinct system for our purposes.

Unfortunately, when providers of new features take language-external approaches, it causes problems for clients. Features cannot be adopted individually, but instead are only available coupled to a collection of other often-unrelated  features (the incidental features included in a newly-designed language or tool, or the existing features of the particular tool that was externally extended). This makes adoption more costly when these other features are either not appropriate or insufficiently developed, or when the features bundled with a different language or tool are simultaneously desirable. For example, although evidence suggests that developers prefer language-integrated parallel programming abstractions to library-based implementations if all else is equal \cite{langvslib}, library-based implementations are more widely adopted because parallel programming languages privilege only a few abstractions, even though parallel programming is rarely the only concern relevant to a component or application. Regular expression support, for example, may be simultaneously desirable when processing large amounts of textual data in parallel. %Similarly, a language and tools designed primarily to support regular expressions might make an interesting research project, but it would not be a suitable tool for writing large applications with more varied needs.

@@

%Clients seem to prioritize the ability to choose different features for different portions of an application. 
%If calling between languages were safe and easy, then using a variety of specialized languages and associated tools might be less problematic. In fact, s
%Recognizing the limitations of relying on monolithic collection of primitives, some researchers have advocated instead for a model where multiple languages used within a single application, calling it the \emph{language-oriented approach} to software development \cite{languageoriented}.

Even when various concerns can be separated into different components, each written in a suitable language (often called the \emph{language-oriented approach} to software development \cite{language-oriented}), the interfaces between these components remains an issue. The specialized primitives particular to one language cannot always be safely and naturally expressed in terms of those available in another, so building a program out of components written in a variety of different languages is difficult or impossible whenever these primitives are exposed at interface boundaries. We refer to this fundamental issue as the \emph{interoperability problem}.

One strategy often taken by programming language designers to partially address the interoperability problem is to  target an established language or bytecode, such as the Java Virtual Machine (JVM) bytecode or LLVM IR, and support a superset of its constructs. Scala \cite{scala} and F\# \cite{fsharp} are two prominent examples of general-purpose languages that have taken this approach, and many domain-specific language frameworks also rely on this strategy (e.g. Delite \cite{delite}). This only enables full interoperability in one direction (Figure \ref{approaches}a). While calling into the common language becomes straightforward, calls in the other direction, or between the languages sharing the common target, are still restricted by the semantics of the common language. 
If new languages only include constructs that can already be expressed safely and reasonably naturally in the common language, then this approach can work well. 
But many of the most innovative constructs found in modern languages (often, the features that justify their creation) are difficult to define in terms of existing constructs in ways that guarantee all necessary invariants are statically maintained and that do not require large amounts boilerplate code and run-time overhead. For example, the type system of F\# guarantees that null values cannot occur within F\# data structures, but maintaining this important invariant still requires run-time checks because the typing rules of F\# do not apply when F\# code is called from other languages on the Common Language Infrastructure (CLI) like C\#. The F\# type system also includes support for checking that units of measure are used correctly \cite{fsharpunits}, but this specialized invariant is left completely unchecked at language boundaries. In Scala, interfaces built around traits that have default method implementations are difficult to implement from Java or other JVM languages and the workaround can break if the trait is modified \cite{scalatraitinterop}. In some cases, desirable features must be omitted entirely due to concerns about  interoperability. The module system in F\#, for example, is substantially simpler than that in its predecessor, OCaml, despite F\# otherwise aiming to maintain compatibility, due to the need for bidirectional interoperability between F\# and other CLI languages.
%\end{itemize}

\subsection{Active Libraries}
For these reasons, we argue that taking a language-external approach to realizing a new feature, by packaging it into a new or derivative language or tool, should be considered harmful and avoided whenever possible. The goal of the research being proposed here is to fundamentally reorganize the core components of the programming system so that such language-external approaches are less frequently necessary, by designing \emph{language-internal extension mechanisms} that give developers control over edit-time and compile-time behaviors that have previously been defined externally and thus, in the case of popular languages, become available only subject to the approval of slow-moving governing committees\footnote{One might compare today's monolithic programming systems to  {centrally-planned} economies, whereas extensible\- systems more closely resemble modern market economies.}\todo{read Arch D. Robison. Impact of economics on compiler optimization.}. Specifically, we will show how control over \textbf{parsing}, \textbf{typechecking}, \textbf{translation} (the first stage of compilation) and \textbf{code completion} can be delegated to user-defined logic distributed in {libraries}, as illustrated in Figure \ref{approaches}b. 
Such libraries are called \emph{active libraries} because, rather than being passive consumers of features already available in the system, they contain logic invoked by the system during development or compilation to provide new domain-specific features  \cite{active-libraries}. Features implemented within active libraries are imported as needed by the clients of libraries that rely on them, unlike features implemented by language-external means.

Some critical issues having to do with {safety} must be overcome before library-based extension mechanisms can be introduced into a programming system, because if too much control over such core aspects of the system is given  to developers, the system may become unreliable. 
%For example, an extension could weaken important metatheoretic guarantees previously provided by the system. 
Type safety, for example, may not hold if the static and dynamic semantics of the language can be modified or extended arbitrarily from within libraries. Furthermore, even if extensions can be shown not to cause such problems in isolation, there may still be conflicts between extensions that could weaken their semantics and lead to subtle problems at link-time. For example, if two active libraries defined differing semantics for the same syntactic form, the issue would only manifest itself when both libraries were imported somewhere within the same program. These kinds of safety issues have plagued previous attempts at language-internal extensibility.% To prevent them, our mechanisms will organize  extension logic around types to guarantee that extensions are both safe in isolation and also safely composable in any combination.

 %This represents a minimalist approach to system design -- the conventional distinction between built-in and user-defined constructs is blurred and most features of the system are orthogonally implemented as {libraries}, rather than by the maintainers of the system.
